\Rpackage{MSnbase} input/output capabilities

### 2013-04-29 inst/doc/MSnbase-io.Rnw
using knitr vignette builder and updates to io

%\VignetteIndexEntry{MSnbase IO capabilities}
%\VignetteKeywords{Mass Spectrometry, Proteomics, Infrastructure }
%\VignettePackage{MSnbase-io}
\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,shadows,fit}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}

@@

\input{Author.tex}

@@

<<knitr, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', 
               par = TRUE,
               prompt = TRUE,
               comment = NA)
options(replace.assign = TRUE, 
        width = 70)
@

@@

\Rpackage{MSnbase}'s aims are to facilitate the reproducible analysis of 
mass spectrometry data within the \R environment, from raw data import and processing, 
feature quantification, quantification and statistical analysis of the results.
Data import functions for several formats are provided and intermediate or final 
results can also be saved or exported. 
These capabilites are presented below.

@@

>>>>

%\VignetteEngine{knitr}
%\VignetteIndexEntry{MSnbase IO capabilities}
%\VignetteKeywords{Mass Spectrometry, Proteomics, Infrastructure }
%\VignettePackage{MSnbase-io}

\documentclass[12pt,a4paper,english]{scrartcl}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}
\usepackage[auth-sc]{authblk}
\usepackage{setspace}
\onehalfspacing

\usetikzlibrary{shapes,arrows,shadows,fit}

% caption formatting
\setcapindent{0em}
\setkomafont{captionlabel}{\sffamily\bfseries}
\setkomafont{caption}{\sffamily}

@@

\author{
  Laurent Gatto\thanks{\email{lg390@cam.ac.uk}}
}

\affil{
  Cambridge Center for Proteomics\\
  University of Cambridge
}

@@

<<knitr, include=FALSE, cache=FALSE>>=
library("knitr")
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRolocdata"))
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', 
               par = TRUE,
               prompt = TRUE,
               comment = NA)
options(replace.assign = TRUE, 
        width = 70)
@

@@

\Rpackage{MSnbase}'s aims are to facilitate the reproducible analysis of 
mass spectrometry data within the \R environment, from raw data import and processing, 
feature quantification, quantification and statistical analysis of the results \citep{Gatto2012}.
Data import functions for several formats are provided and intermediate or final 
results can also be saved or exported. 
These capabilites are presented below.

@@

\section{Creating Robject{MSnSet} from text spread sheets}

This section describes the generation of \Robject{MSnSet} objects using 
data available in a text-based spreadsheet. This entry point into \R and 
\Rpackage{MSnbase} allows to import data processed by any of the 
third party mass-spectrometry processing software available and proceed 
with data exploration, normalisation and statistical analysis using 
functions available in \R and the numerous Bioconductor packages.

\subsection{A complete workflow}

The following section describes a workflow that uses three input files to 
create the \Robject{MSnSet}. These files respectively describe the quantitative 
expression data, the sample meta-data and the feature meta-data. 
It is taken from the \Rpackage{pRolco} tutorial and uses example files from 
the \Rpackage{pRolocdata} package.

\subsubsection{The original data file}\label{sec:orgcsv}

We start by describing the  \texttt{csv} to be used as input using the \Rfunction{read.csv} function.

<<readCsvData0>>=
## The original data for replicate 1, available
## from the pRolocdata package
f0 <- dir(system.file("extdata", package = "pRolocdata"), 
          full.names = TRUE, pattern = "pr800866n_si_004-rep1.csv")
csv <- read.csv(f0)
@

The three first lines of the original spreadsheet, containing the data for replicate one, are illustrated below (using the function \Rfunction{head}). It contains \Sexpr{nrow(csv)} rows (proteins) and \Sexpr{ncol(csv)} columns, including protein identifiers, database accession numbers, gene symbols, reporter ion quantitation values, information related to protein identification, \ldots

<<showOrgCsv>>=
head(csv, n=3)
@

Below read in turn the spread sheets that contain the quantitation data (\texttt{exprsFile.csv}),
feature meta-data (\texttt{fdataFile.csv}) and sample meta-data (\texttt{pdataFile.csv}).

<<readCsvData1, tidy = FALSE>>=
## The quantitation data, from the original data
f1 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "exprsFile.csv")
exprsCsv <- read.csv(f1)
## Feature meta-data, from the original data
f2 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "fdataFile.csv")
fdataCsv <- read.csv(f2)
## Sample meta-data, a new file
f3 <- dir(system.file("extdata", package = "pRoloc"), 
          full.names = TRUE, pattern = "pdataFile.csv")
pdataCsv <- read.csv(f3)
@

\begin{description}

\item[\texttt{exprsFile.csv}] containing the quantitation (expression) data for the \Sexpr{nrow(exprsCsv)} proteins and 4 reporter tags.
<<showExprsFile>>=
head(exprsCsv, n=3)
@
\item[\texttt{fdataFile.csv}] containing meta-data for the \Sexpr{nrow(fdataCsv)} features (here proteins).
<<showFdFile>>=
head(fdataCsv, n=3)
@
\item[\texttt{pdataFile.csv}] containing samples (here fractions) meta-data. This simple file has been created manually.
<<showPdFile>>=
pdataCsv
@
\end{description}

The self-contained \Robject{MSnSet} can now easily be generated using the \Rfunction{readMSnSet} constructor, 
providing the respective \texttt{csv} file names shown above and specifying that the data is comma-separated 
(with \Robject{sep = ","}). Below, we call that object \Robject{res} and display its content.

<<makeMSnSet, tidy = FALSE>>=
library("MSnbase")
res <- readMSnSet(exprsFile = f1,
                  featureDataFile = f2,
                  phenoDataFile = f3,
                  sep = ",")
res
@

\subsubsection{The \Robject{MSnSet} class}

Although there are additional specific sub-containers for additional meta-data 
(for instance to make the object MIAPE compliant), the feature 
(the sub-container, or slot \Robject{featureData}) and sample (the \Robject{phenoData} slot) 
are the most important ones. They need to meet the following validity requirements 
(see figure \ref{fig:msnset}):

\begin{itemize}
\item the number of row in the expression/quantitation data and feature data must 
  be equal and the row names must match exactly, and  
\item the number of columns in the expression/quantitation data and number of row 
  in the sample meta-data must be equal and the column/row names must match exactly. 
\end{itemize}

A detailed description of the \Robject{MSnSet} class is available by typing \Rfunction{?MSnSet} in the \R console.

\begin{figure}[!hbt]
\centering
    \includegraphics[width=0.5\textwidth]{msnset.png}
\caption{Dimension requirements for the respective expression, feature and sample meta-data slots. }
\label{fig:msnset}
\end{figure}

The individual parts of this data object can be accessed with their respective accessor methods:

\begin{itemize}
\item the quantitation data can be retrieved with \Rfunction{exprs(res)}, 
\item the feature meta-data with \Rfunction{fData(res)} and 
\item the sample meta-data with \Rfunction{pData(res)}. 
\end{itemize}
